---
title: "Racial/Ethnic Disparities in Survival among Young Patients with Non-Small Cell Lung Cancer Patients: A SEER Analysis"
author: "Carissa Feliciano"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
---

# Introduction

Lung cancer is the leading cause of cancer deaths for both men and women in the US, with non-small cell lung cancer (NSCLC) accounting for 80-85% of cases. While several studies have investigated racial/ethnic disparities among all lung cancer patients, there is limited data characterizing racial/ethnic disparities among young NSCLC patients (aged ≤50 years). This project evaluated whether survival time was associated with race/ethnicity among among young adults (aged 18-50 years) with non-small cell lung cancer (NSCLC). Additionally, this project aimed to determine whether any differences in survival across the racial/ethnic groups could be attributed to differences in the frequency distributions of sex, histological subtype, and stage at diagnosis. This project utilized the Surveillance, Epidemiology, and End Results (SEER) 17 Database and included patients aged 18 to 50 years with an incident diagnosis of non-small cell lung cancer between January 1, 2011 and December 31, 2021, first primary malignancy, and residence in California (n = 4427).

The full report can be downloaded [here](https://github.com/carissafeliciano/PM566-final-project/blob/master/docs/Final_CarissaFeliciano.pdf).

Instructions on how to access the SEER-17 dataset can be found [here](https://github.com/carissafeliciano/PM566-final-project/blob/master/data/readme.md).

# Interactive Visualizations

```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
library(ggplot2)
library(plotly)
library(dplyr)
library(DT)
library(knitr)
library(boot)
```

```{r, echo=FALSE}
# Initialize code chunk options
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = TRUE,
  cache = FALSE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px")
```

```{r load-data, echo=FALSE}
source("process_SEER_data.R")
```

```{r, include=FALSE, echo=FALSE}
source("process_SEER_data.R")
p1_hist <- ggplot(seer, aes(x = survival, fill = race)) +
   geom_histogram(alpha = 0.7, position = "identity", bins = 30) +
  labs(
    title = "Histogram of Survival Time by Race/Ethnicity",
    x = "Survival Time (Months)",
    y = "Count",
    fill = "Race/Ethnicity"
  ) +
  theme_minimal()

ggplotly(p1_hist)
```

```{r, echo=FALSE}
# Boxplot of survival time by race/ethnicity 

race_xaxis <- c(
  "Non-Hispanic White" = "NHW",
  "Non-Hispanic Black" = "NHB",
  "Hispanic (All Races)" = "Hispanic",
  "Non-Hispanic Asian or Pacific Islander" = "NHAPI",
  "Non-Hispanic American Indian/Alaska Native" = "NHAIAN"
)

p1_boxplot <- ggplot(seer) +
  geom_boxplot(mapping = aes(x = race, y = survival, fill = race)) +
  labs(
    title = "Figure 1: Distribution of Survival Times by Race/Ethnicity",
    x = "Race/Ethnicity", 
    y = "Survival Time (Months)",
    fill = "Race/Ethnicity"
    )+
  scale_x_discrete(labels = race_xaxis) +
  theme_minimal()+ theme(axis.line=element_line(color = "black")) +
  theme(
    axis.text.x = element_text(size = 9),
    axis.text.y = element_text(size = 9),
    axis.title = element_text(size = 10),
    axis.title.x = element_text(margin = margin(t = 12)),
    axis.title.y = element_text(margin = margin(r = 12)),
    plot.title = element_text(hjust = 0, face = "bold", size = 12)
  )

p1_ggplotly <- ggplotly(p1_boxplot) |>
  layout(
    autosize = TRUE,
    margin = list(t = 100)  # Adjust top margin
  )

p1_ggplotly
```

<br>

The boxplots display the distribution of survival times for each racial/ethnic group. The line within each box represents the median survival time. The upper bound of the box represents the 75th percentile (Q3), and the lower bound of the box represents the 25th percentile (Q1). The upper whisker is the maximum (highest value within 1.5\*IQR of Q3), and the lower whisker is the minimum (lowest value within 1.5\*IQR of Q1). Points outside the whiskers represent outliers. NHW, non-Hispanic White; NHB, non-Hispanic Black; Hispanic, Hispanic all races; NHAPI, non-Hispanic Asian or Pacific Islander; NHAIAN, non-Hispanic American Indian/Alaska Native.

The median survival time was lowest for the NHB group (11.0 months) and highest for the NHAIAN group (21.5 months). The median survival time was similar for the NHW (19 months), Hispanic (18 months), and NHAPI (19 months) groups.

```{r, echo=FALSE}
# Barchart of median survival time with 95% CI by race/ethnicity and sex
# Function to compute median and 95% CI using bootstrapping
median_ci <- function(data, indices) {
  boot_data <- data[indices]
  med <- median(boot_data)
  return(med)
}

# Calculate median survival time
# Perform bootstrapping to get 95% CI for median
survival_race_sex <- seer |>
  group_by(race, sex) |>
  summarise(
    med_survival = median(survival),
    boot_obj = list(boot(survival, median_ci, R = 1000)),
    lower_ci = boot.ci(boot_obj[[1]], type = "perc")$percent[4],  # 2.5th percentile
    upper_ci = boot.ci(boot_obj[[1]], type = "perc")$percent[5]   # 97.5th percentile
  )

survival_race_sex$hover_text <- paste(
  "Race: ", survival_race_sex$race, 
  "<br>Sex: ", survival_race_sex$sex,
  "<br>Median Survival: ", survival_race_sex$med_survival, "months", 
  "<br>95% CI: ", sprintf("%.2f", survival_race_sex$lower_ci), "-", sprintf("%.2f", survival_race_sex$upper_ci)
)

p2_bar_survival_race_sex <- ggplot(survival_race_sex) +
  geom_bar(
      mapping = aes(x = race, y = med_survival, fill = sex, text = hover_text), 
      stat = "identity", 
      position = position_dodge(width = 0.9),
      color = "black"
    ) +
  geom_errorbar(
      mapping = aes(x = race, ymin = lower_ci, ymax = upper_ci, group = sex, fill = sex),
      position = position_dodge(width = 0.9), 
      width = 0.25
    ) +
  labs(
      title = "Figure 2: Median Survival Time by Race/Ethnicity and Sex",
      x = "Race/Ethnicity", 
      y = "Median Survival Time (Months)",
      fill = "Sex"
    ) +
  scale_fill_manual(values = c("Male" = "cornflowerblue", "Female" = "lightcoral")) +
  scale_x_discrete(labels = race_xaxis) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_minimal()+ theme(axis.line=element_line(color = "black")) +
  theme(
    axis.text.x = element_text(size = 9),
    axis.title = element_text(size = 10),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    legend.position = "bottom",
    legend.background = element_rect(color = "black", fill = NA),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.key.size = unit(0.4, "cm"),
    plot.title = element_text(hjust = 0, face = "bold", size = 12)
    ) 

ggplotly(p2_bar_survival_race_sex, tooltip = "text")
```

<br>

This figure displays the median survival time by race/ethnicity and sex. Error bars represent 95% confidence intervals (CIs). NHW, non-Hispanic White; NHB, non-Hispanic Black; Hispanic, Hispanic all races; NHAPI, non-Hispanic Asian or Pacific Islander; NHAIAN, non-Hispanic American Indian/Alaska Native.

The median survival time of NHB females (12 months) was significantly lower than that of NHW (21 months), Hispanic (21 months), and NHAPI (25 months) females. Similarly, the median survival time of NHB males (10 months) was significantly lower than that of NHW (17 months), Hispanic (14 months), and NHAPI (15 months) males.

```{r, echo=FALSE}
#Barchart of median survival time with 95% CI by race/ethnicity and stage at diagnosis

survival_race_stage <- seer |>
  filter(stage != "Unknown/unstaged") |>
  group_by(race, stage) |>
  summarise(
    med_survival = median(survival),
    boot_obj = list(boot(survival, median_ci, R = 1000)),
    lower_ci = boot.ci(boot_obj[[1]], type = "perc")$percent[4],  # 2.5th percentile
    upper_ci = boot.ci(boot_obj[[1]], type = "perc")$percent[5]   # 97.5th percentile
  )

survival_race_stage$hover_text <- paste(
  "Race: ", survival_race_stage$race, 
  "<br>Stage: ", survival_race_stage$stage,
  "<br>Median Survival: ", survival_race_stage$med_survival, "months", 
  "<br>95% CI: ", sprintf("%.2f", survival_race_stage$lower_ci), "-", sprintf("%.2f", survival_race_stage$upper_ci)
)

p3_bar_survival_race_stage <- ggplot(survival_race_stage) +
  geom_bar(mapping = aes(x = race, y = med_survival, fill = stage, text = hover_text), 
           stat = "identity", 
           position = position_dodge(width = 0.9), 
           color = "black") +
  geom_errorbar(
      mapping = aes(x = race, ymin = lower_ci, ymax = upper_ci, group = stage, fill = stage),
      position = position_dodge(width = 0.9), 
      width = 0.25
    ) +
  scale_fill_brewer(palette = "YlOrBr") +
  labs(
    title = "Figure 3: Median Survival Time by Race/Ethnicity and Stage at Diagnosis",
    x = "Race/Ethnicity", 
    y = "Median Survival Time (Months)",
    fill = "Stage at Diagnosis") +
  scale_x_discrete(labels = race_xaxis) +
  scale_y_continuous(
    limits = c(0,85),
    expand = expansion(mult = c(0, 0.05))) +
  theme_minimal()+ theme(axis.line=element_line(color = "black")) +
  theme(
    axis.text.x = element_text(size = 9),
    axis.title = element_text(size = 10),
    axis.title.x = element_text(margin = margin(t = 8)),
    axis.title.y = element_text(margin = margin(r = 8)),
    legend.background = element_rect(color = NA, fill = NA),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.key.size = unit(0.4, "cm"),
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0, face = "bold", size = 12)
    ) 

ggplotly(p3_bar_survival_race_stage, tooltip = "text")
```

<br>

This figure displays the median survival time by race/ethnicity and stage at diagnosis. Error bars represent 95% CIs. NHW, non-Hispanic White; NHB, non-Hispanic Black; Hispanic, Hispanic all races; NHAPI, non-Hispanic Asian or Pacific Islander; NHAIAN, non-Hispanic American Indian/Alaska Native.

Among patients with localized cancer at diagnosis, NHB patients had a lower median survival time (43.5 months) than NHW (57.5 months), NHAPI (51 months), and NHAIAN (75 months) patients. However, these differences were not statistically significant as the 95% CIs overlapped. The median survival time for Hispanic patients (40.5 months) was similar to that of NHB patients (43.5 months). Among patients with regional cancer at diagnosis, NHB patients had a lower median survival time (23 months) than NHW (35.5 months), Hispanic (31 months), and NHAPI (45 months) patients. These differences were not statistically significant. Among patients with distant cancer at diagnosis, the median survival time of NHB patients (6 months) was significantly lower than that of NHW (10 months), Hispanic (12 months), and NHAPI (16 months) patients.

```{r, echo=FALSE}
#Barchart of median survival time with 95% CI by race/ethnicity and histologic diagnosis
survival_race_hist <- seer |>
  group_by(race, histology) |>
  summarise(
    med_survival = median(survival),
    boot_obj = list(boot(survival, median_ci, R = 1000)),
    lower_ci = boot.ci(boot_obj[[1]], type = "perc")$percent[4],  # 2.5th percentile
    upper_ci = boot.ci(boot_obj[[1]], type = "perc")$percent[5]   # 97.5th percentile
  )

survival_race_hist$hover_text <- paste(
  "Race: ", survival_race_hist$race, 
  "<br>Histology: ", survival_race_hist$histology,
  "<br>Median Survival: ", survival_race_hist$med_survival, "months", 
  "<br>95% CI: ", sprintf("%.2f", survival_race_hist$lower_ci), "-", sprintf("%.2f", survival_race_hist$upper_ci)
)

p4_bar_survival_race_hist <- ggplot(survival_race_hist) +
  geom_bar(mapping = aes(x = race, y = med_survival, fill = histology, text = hover_text), 
           stat = "identity", 
           position = position_dodge(width = 0.9), 
           color = "black") +
  geom_errorbar(
      mapping = aes(x = race, ymin = lower_ci, ymax = upper_ci, group = histology, fill = histology),
      position = position_dodge(width = 0.9), 
      width = 0.25
    ) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Figure 4: Median Survival Time by Race/Ethnicity and Histological Diagnosis",
    x = "Race/Ethnicity", 
    y = "Median Survival Time (Months)",
    fill = "Histological Subtype") +
  scale_x_discrete(labels = race_xaxis) +
  scale_y_continuous(
    limits = c(0,70),
    expand = expansion(mult = c(0, 0.05))) +
  theme_minimal()+ theme(axis.line=element_line(color = "black")) +
  theme(
    axis.text.x = element_text(size = 9),
    axis.title = element_text(size = 10),
    axis.title.x = element_text(margin = margin(t = 8)),
    axis.title.y = element_text(margin = margin(r = 8)),
    legend.background = element_rect(color = NA, fill = NA),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.key.size = unit(0.4, "cm"),
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0, face = "bold", size = 11)
    )

ggplotly(p4_bar_survival_race_hist, tooltip = "text")
```

<br>

This figure displays the median survival time by race/ethnicity and histological diagnosis. Error bars represent 95% CIs. NHW, non-Hispanic White; NHB, non-Hispanic Black; Hispanic, Hispanic all races; NHAPI, non-Hispanic Asian or Pacific Islander; NHAIAN, non-Hispanic American Indian/Alaska Native.

Among patients diagnosed with adenocarcinoma, NHB patients had a statistically significantly lower median survival (11 months) than NHW (16 months), Hispanic (16 months), and NHAPI (21 months) patients. Among patients diagnosed with squamous cell carcinoma, NHB patients had a lower median survival (10 months) than NHW (13.5 months) and NHAPI (11.5 months). However, this difference was not statistically significant. Among patients with a histological diagnosis of NOS, NHB patients (12.5 months) had a statistically significantly lower median survival than NHW (37.5 months) and Hispanic (30 months). NHB patients had a lower median survival (12.5 months) than NHAPI patients (18.5 months) as well, but this difference appeared to be not statistically significant. Less than 3% of patients from each group were diagnosed with large cell carcinoma. The small sample size of this subgroup limits the ability to draw definitive conclusions, as reflected by the wide 95% CIs.
